---
title: "Assignment 1"
subtitle: "Introduktion til Sundhedsdata · Worked on in lessons 2–3 · Presented lesson 4"
author: "Name: ____________"
date: last-modified
lang: en
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## The task

In this assignment, you will be doing some basic data manipulation to set up a dataset for analysis. You will be using the `.dta` files in this repository, and your outputs will be a single merged dataset.

## How to work on this file

This file is a partly complete data management script. 

Nearly everything you need is in the demo, `demo_part1.qmd`. Each step below
names the section of the demo that covers it. 

## Setup

```{r}
#| label: setup
library(tidyverse)
library(haven)

raw_data_dir <- "data/raw/"
processed_data_dir <- "data/processed/202609/"
```

## Tasks

Outcome: a dataset with one row per person, for the year 2019. All Danish individuals ages 18 to 65 who are employed.

Your final dataset must use exactly these column names, so that we can all run the same checks on each other's work in lesson 4:

| Column | Type | What it holds |
|---|---|---|
| `pnr` | character | Person identifier |
| `alder` | numeric | Age in years |
| `koen` | factor | Sex, in human readable words |
| `civ` | factor | Marital status, in human readable words |
| `uddannelse` | factor | Highest achieved education level, in human readable words |
| `indkomst` | numeric | Income level |
| `gen_indkomst` | numeric | Average income of individuals with the same education level |
| `forskel_indkomst` | numeric | Difference between the individual's income and the average for their education level |
| `uddannelse_alder` | numeric | Age at which the education level was achieved |

No other columns included.

Also submit a data dictionary, which describes the contents of your dataset and how it was derived.

### 1. Read in data and take a look at the structure of variables
5 data sources: BEF, UDDA, IND, and the HFAUDD and PRE_SOCIO text files

*Demo: "Reading a register file", "Looking at what you have", "Checking the granularity".*

```{r}
# TODO: read in bef2019.dta, uddf2024.dta, ind2019.dta and hfaudd_disced15.txt,
# and look at the structure and contents of each one.

```

Freebie: pre_socio is a little complicated to parse, so it is provided here
```{r}
pre_socio <- read_delim(paste0(raw_data_dir,"pre_socio_klassifikation.txt"), delim = ";")

pre_socio <- pre_socio %>%
  select(pre_socio = START, arbejdsmarked = ANALYSE_KAT) %>%
  mutate(pre_socio = as.character(pre_socio))
```

### 2. Managing categorical variables and variable types
Three steps:
  1. Select the columns you need
  2. Filtering to the rows you want
  3. Recoding categorical columns (where it's simple - leave out `hfaudd` for now until merging)

*Demo: "Recoding a categorical variable", "Selecting rows and columns", "Classification keys".*

```{r}
# TODO: cut each dataset down to the columns and rows you need, and recode the
# categorical variables that are straightforward to recode.

```

### 3. Merging all datasets together

Watch the row count - make sure that you're not unexpectedly dropping data, or 
creating missing values.

*Demo: "Attaching the key to the data", "Joining two registers", "Choosing the type of join".*

```{r}
# TODO: merge your datasets into a single dataset called dt.

```

Check: depending on the type of merge you do, you may end up with some non-employed
individuals in your sample. Double check that this is not the case:

```{r}
count(dt, arbejdsmarked)
```

```{r}
# TODO: if the check above shows people who are not employed, drop them.

```

Note that you may need to change the dataset name or variable name to get the test to work.

### 4. Create new variables

Three of the columns in the table above do not exist in any of the register
files — you have to build them.

Hint: to create the "average income level at each education level" variable, you'll need to do a group_by() and aggregation.

*Demo: "Creating a new column", "Aggregation and summarisation".*

```{r}
# TODO: create uddannelse_alder, gen_indkomst and forskel_indkomst.

```

### 5. Checks and selecting final columns
Check whether the education levels have any missingness:

```{r}
count(dt, uddannelse)
```

If missingness exists, recode it to "Unknown".

```{r}
# TODO: recode missing education levels, if there are any.

```

```{r}
# TODO: keep only the columns listed in the table at the top of this file, with
# exactly those names, then look at the finished dataset with glimpse() and
# summary().

```

### 6. Writing the dataset

*Demo: "Classification keys" (for `write_csv`), "Documenting the data: the data dictionary".*

```{r}
# TODO: write your finished dataset to processed_data_dir.

```

Your data dictionary describes what is in the dataset: one row per
column, with the column name, its type, the register it came from, and a
description of how it was made. Include the population definition at the top
of the dictionary as well. See the demo for the version we use in this
course.

```{r}
# TODO: your data dictionary.

```

## Extra - challenges

These are optional. If you solve one, you may be asked to present it in lesson 4.

### Challenge 1: something is wrong with the education variable

The education level in your dataset is not correct for everybody. For some of
the people in your data, the education level you have assigned them is not one
they actually held in 2019.

Work out who is affected and why, fix it, and be ready to explain what went
wrong. When you have found the problem, check whether any *other* column in
your dataset is wrong for the same reason.

```{r}

```

### Challenge 2: process more than one year

`data/raw/` has BEF and IND for 2019 through 2024, not just 2019. Build the
same analysis file for every year at once, without copying and pasting your
code six times, and end up with a single dataset that has one row per person
per year.

*Demo: "Functions and loops".*

```{r}

```

## Submission and presentation

### What to submit

Please submit your work as **two files**, to the Assignment 1 folder on
itslearning:

1. `assignment1_efternavn.qmd` — your completed script, with your surname in
   the filename.
2. `assignment1_efternavn.html` — the rendered output, which is how we can verify that your code runs

If you wrote your data dictionary as a separate file, please also include this file in your upload.

**Deadline: Tuesday 22 September, 9:00** — before lesson 4.
