---
title: "Assignment 2"
subtitle: "Introduktion til Sundhedsdata · Worked on in lessons 5–7 · Presented lesson 8"
author: "Name: ____________"
date: last-modified
lang: en
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  eval: true
  message: false
---

## How to work on this file

This file is a partly complete analysis script. Each task asks for one or more plots, and names the section of the demo, `demo_part2.qmd`, that covers it. Nearly everything you need is in the demo.

Work through the tasks in order, since later tasks reuse data created in earlier ones. Render the file often to check that your code runs successfully.

Every plot should be presentable. Give each a title, readable axis labels and legend titles, and choose colors that make the comparison clear.

## Setup

```{r}
#| label: setup
library(tidyverse)
library(arrow)
library(ggExtra)
library(ggcorrplot)
library(knitr)

processed_data_dir <- "data/processed/202609/"

expbod <- open_dataset(paste0(processed_data_dir, "expbod_teaching_parquet/")) |> collect()
health_spend <- read_parquet(paste0(processed_data_dir,"SyntheticData_healthcareExp_100k_251029.parquet"))

health_spend <- health_spend |> 
    mutate(education = case_when(
        education_long_higher == 1 ~ "Long higher",
        education_medium_long == 1 ~ "Medium-long",
        education_upper_secondary == 1 ~ "Upper secondary",
        education_primary == 1 ~ "Primary",
        education_other == 1 ~ "Other",
        .default = "Missing"
    )) |> 
    mutate(sex = factor(sex, labels = c("Male","Female")))

health_spend <- health_spend |> 
    mutate(education = factor(education, levels = c("Primary","Upper secondary","Medium-long","Long higher","Other","Missing")))
```


### 1. Univariate plots

**Task 1:** plot the distribution of wealth and income in the data, stratified by sex and education

Note that you may want to consider using a logarithmic scale, but first check to see if this switch would remove some rows of data.

*Demo: "Univariate plots".*

```{r}
# TODO: plot the distribution of wealth and income, stratified by sex and education.

```


**Task 2:** plot the distribution of healthcare costs across rows

*Demo: "Univariate plots" (faceting).*

```{r}
# TODO: pivot the cost columns to long format, and plot the distribution of each type of spending.

```

::: {.callout-tip title="Checkpoint"}
After pivoting, your data should have one row per person per type of spending: 800,000 rows (100,000 people × 8 types of spending), or 600,000 rows if you drop the two psychiatric types.
:::

Are there types of spending that have no non-zero cost amounts? Is there a way to remove these from the visualization?

Extra: see if you can aggregate and plot the distribution of total spending
```{r}
# TODO: aggregate spending to one row per person, and plot the distribution of total spending.

```

::: {.callout-tip title="Checkpoint"}
Your aggregated data should have one row per person: 100,000 rows. 101 people have zero total spending.
:::

**Task 3:** plot the frequency of ICD-10 chapters by education level

*Demo: "Univariate plots" (bar charts).*

```{r}
# TODO: pivot the ICD-10 chapter columns to long format, and plot the frequency of each chapter by education level.

```

::: {.callout-tip title="Checkpoint"}
After pivoting and keeping only diagnoses, your data should have one row per person per chapter they have a diagnosis in: 109,844 rows, covering 25 chapters and 51,885 people.
:::

Note that a person with diagnoses in several chapters appears in several bars.

### 2. Bivariate plots

**Task 4:** plot inpatient spending against prescription drug spending

*Demo: "Multivariate plots".*

```{r}
# TODO: plot inpatient spending against prescription drug spending.

```

Most people have no inpatient spending. How many people have both inpatient and prescription drug spending above zero? Plot only these people, on a logarithmic scale:

```{r}
# TODO: keep only people with both types of spending above zero, count them, and plot them on a logarithmic scale.

```

**Task 5:** plot total health spending against income, with the distribution of each variable on the sides of the plot

*Demo: "Multivariate plots" (marginal distributions).*

```{r}
# TODO: join total spending onto health_spend, and plot it against income with the distribution of each variable on the sides.

```

::: {.callout-tip title="Checkpoint"}
After joining total spending onto `health_spend`, it should still have 100,000 rows. If it has more, the join is duplicating people.
:::

**Task 6:** make a correlation plot of age, income, wealth and the spending variables

*Demo: "Heatmaps" (correlation plots).*

```{r}
# TODO: make a correlation matrix with cor(), and plot it with ggcorrplot().

```

::: {.callout-tip title="Checkpoint"}
Your correlation matrix should have one row and one column per variable: 10 × 10.
:::

Hint: the psychiatric variables are all zero, so leave them out of the correlation plot.

Do the relationships between these variables make sense to you? Give an explanation here to 
describe which correlations can be explained and which are surprising.

*YOUR DESCRIPTION HERE*


**Task 7:** plot the distribution of total spending by ICD-10 chapter, and then average spending by ICD-10 chapter and type of spending

*Demo: "Univariate plots" (boxplots, bar charts with `geom_col()`).*

```{r}
# TODO: plot the distribution of total spending by ICD-10 chapter, then calculate and plot average spending by chapter and type of spending.

```

::: {.callout-tip title="Checkpoint"}
Joining total spending onto the ICD-10 data should not change the number of rows: still 109,844. The averages by chapter and type of spending should have one row per chapter per type: 200 rows (25 chapters × 8 types of spending), or 150 rows if you drop the two psychiatric types.
:::

### 3. Tables

**Task 8:** make a Table 1 describing the individuals in the data by education level

*Demo: "Tables".*

```{r}
# TODO: make a Table 1 describing the individuals in the data by education level.

```

::: {.callout-tip title="Checkpoint"}
Your table should have one row per characteristic and one column per education level (6), and the N row should sum to 100,000.
:::

## Open-ended hypothesis testing
The other tasks in this assignment have been specific about what you should produce. However, there are many 
relationships in both datasets which have not been explored yet. Use the demo, the other assignment tasks, and 
the online resources to build **2 plots** which examine other questions you may have in either dataset.

```{r}

```



## Submission and presentation

### What to submit

Please submit your work as **two files**, to the Assignment 2 folder on itslearning:

1. `assignment2_efternavn.qmd` — your completed script, with your surname in the filename.
2. `assignment2_efternavn.html` — the rendered output, which is how we can verify that your code runs.

**Deadline: TODO** — before lesson 8.
